Scaling up the hyperparameter grid for Naomi

### 2023-03-27 src/dev_scale-grid/scale-grid.Rmd
Clarify writing of scale-grid, and correct calculation of Gaussian integral #22

  **Task** We would like to begin to integrate over the hyperparameters using some kind of grid structure. The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical. In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
---

@@

Start by fitting the model with `TMB`.
This is quick to do and allows us to get a model we can play with to figure out which hyperparmeters are important.

@@

# Grid ideas

@@

There are `r length(hypers)` of them, if you'd like to count.

@@

## Number of points proportional to standard deviation

@@

This would look like the following:

@@

A limitation of this approach is that some hyperparameters have higher SD than others in part due to the different scales that they are on.
For example if the hyper is on the logit scale then (from the plot above) it looks as though it will have a higher SD than another hyper on the log scale.
We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension: doing this in any good automated ways would be a nice contribution.

## Eigendecomposition

Another option is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.

@@

So with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.

What do the PC loadings look like?

@@

C_plot

C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C_s[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank()
  )

C_plot + C_s_plot
```

@@

### Testing with simple examples

@@

obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov)
}

@@

ground_truth <- cbind(grid, pdf = obj(grid))

@@

gaussian_2d <- pca_rescale(m = mu, C = cov, s = 1, k = 5)

plot +
  geom_point(
    data = data.frame(mvQuad::getNodes(gaussian_2d)) %>%
      mutate(weight = mvQuad::getWeights(gaussian_2d)),
    aes(x = X1, y = X2, size = weight),
    col = multi.utils::cbpalette()[2],
    inherit.aes = FALSE
  ) +
  scale_size_continuous(range = c(1, 2))
```

Test how this PCA grid works for some simple integrals.
We will use Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.

@@

truemode2d <- c(log((sum(y1) + 1)/(length(y1) + 1)), log((sum(y2) + 1)/(length(y2) + 1)))

truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
truelognormconst2d <- truelogint(y1) + truelogint(y2)

objfunc2d <- function(x) logfteta2d(x, c(y1, y2))

funlist2d <- list(
  fn = objfunc2d,
  gr = function(x) numDeriv::grad(objfunc2d,x),
  he = function(x) numDeriv::hessian(objfunc2d,x)
)

opt_bfgs_2d <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))

norm_bfgs_2d_1 <- aghq::normalize_logpost(opt_bfgs_2d, 1, 1)
norm_bfgs_2d_3 <- aghq::normalize_logpost(opt_bfgs_2d, 3, 1)
norm_bfgs_2d_5 <- aghq::normalize_logpost(opt_bfgs_2d, 5, 1)
norm_bfgs_2d_7 <- aghq::normalize_logpost(opt_bfgs_2d, 7, 1)

pca_grid_bfgs_2d_1 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 1)
pca_grid_bfgs_2d_3 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 3)
pca_grid_bfgs_2d_5 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 5)
pca_grid_bfgs_2d_7 <- pca_rescale(m = opt_bfgs_2d$mode, C = solve(opt_bfgs_2d$hessian), s = 1, k = 7)

plot_points <- function(gg) {
  gg$nodes %>%
    as.data.frame() %>%
    mutate(weights = gg$weights) %>%
    ggplot() +
    geom_point(
      aes(x = V1, y = V2, size = weights),
      alpha = 0.8,
      col = multi.utils::cbpalette()[2],
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2)) +
    labs(x = "", y = "", size = "Weight") +
    theme_minimal()
}

{plot_points(pca_grid_bfgs_2d_1) + plot_points(pca_grid_bfgs_2d_3)} / {plot_points(pca_grid_bfgs_2d_5) + plot_points(pca_grid_bfgs_2d_7)}

norm_pca_bfgs_2d_1 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_1, adapt = FALSE)
norm_pca_bfgs_2d_3 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_3, adapt = FALSE)
norm_pca_bfgs_2d_5 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_5, adapt = FALSE)
norm_pca_bfgs_2d_7 <- local_normalize_logpost(opt_bfgs_2d, basegrid = pca_grid_bfgs_2d_7, adapt = FALSE)

results_2d <- data.frame(
  "method" = c(
    "Truth",
    paste0("aghq, k = ", c(1, 3, 5, 7)),
    paste0("aghq-pca, k = ", c(1, 3, 5, 7))
  ),
  "lognormconst" = c(
    truelognormconst2d,
    norm_bfgs_2d_1$lognormconst, norm_bfgs_2d_3$lognormconst, norm_bfgs_2d_5$lognormconst, norm_bfgs_2d_7$lognormconst,
    norm_pca_bfgs_2d_1$lognormconst, norm_pca_bfgs_2d_3$lognormconst, norm_pca_bfgs_2d_5$lognormconst, norm_pca_bfgs_2d_7$lognormconst
  )
)

ggplot(results_2d, aes(x = method, y = lognormconst)) +
  geom_point(size = 3) +
  labs(x = "Method", y = "Log normalising constant") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
```

### Application to Naomi

>>>>

  **Task** We would like to begin to integrate over the hyperparameters using multiple nodes. The problem is that there are >20 hyperparameters, and so using any type of dense grid is not practical. In this notebook we investigate approaches to using a smaller number of grid points to cover the space, including 1) allocating points in proporiton to posterior standard deviation 2) using principle components analysis.
---

@@

Start by fitting the model with `TMB`.
This is quick to do and allows us to get a model we can play with to figure out which hyperparmeters are important.
In particular we obtain the mode and Hessian at the mode of the Laplace approximation to the hyperparameter marginal posterior.

@@

# Number of points proportional to standard deviation

@@

There are `r length(hypers)` of them, if you'd like to count.
There are this many on the logit scale:

```{r}
data.frame(hypers) %>%
  filter(str_detect(hypers, "^logit_")) %>%
  nrow()
```

And this many on the log scale:
```{r}
data.frame(hypers) %>%
  filter(str_detect(hypers, "^log_")) %>%
  nrow()
```

@@

@@

This would look as follows:

@@

A limitation of this approach is that some hyperparameters have higher SD than others in part due to the different scales that they are on.
For example if the hyper is on the logit scale then it looks as though it will have a higher SD than another hyper on the log scale based on the plot above.
To be more precise:

```{r}
data.frame(par = names(sd), sd = unname(sd)) %>%
  filter(str_detect(par, "^log_")) %>%
  summarise(mean_sd = mean(sd))
```

```{r}
data.frame(par = names(sd), sd = unname(sd)) %>%
  filter(str_detect(par, "^logit_")) %>%
  summarise(mean_sd = mean(sd))
```

Does this really reflect our intuititions about what is important to take greater account of?
We would ideally like to know how much variation in each hyperparameter effects variation in model outputs, and then take that into account in the importance we place on each dimension: doing this in any good automated ways would be a nice contribution.

# Eigendecomposition

Although it doesn't solve the above issue, another option to using the standard deviations is to use principle components analysis (PCA) to create a grid on a subspace of $\mathbb{R}^{24}$ which retains most of the variation.

@@

So with 5 dimensions included, we can explain `r round(100 * filter(tv_df, n == 5) %>% pull(tv), 1)`% of the total variation.
Or, with 10 dimensions included, that percentage increases to `r round(100 * filter(tv_df, n == 10) %>% pull(tv), 1)`%.
What do the PC loadings (columns of the matrix `E`) look like?

@@

C_s_plot <- reshape2::melt(as.matrix(C_s)) %>%
  ggplot(aes(x = Var1, y = Var2, fill = value)) +
  labs(x = "", y = "", fill = "C_s[i, j]") +
  geom_tile() +
  scale_fill_viridis_c() +
  labs(x = "i", y = "j") +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "bottom"
  )

{C_plot + theme(legend.position = "bottom")} + C_s_plot
```

@@

## Testing with simple examples

### Gaussian

@@

obj <- function(theta) {
  mvtnorm::dmvnorm(theta, mean = mu, sigma = cov, log = TRUE)
}

ff <- list(
  fn = obj,
  gr = function(theta) numDeriv::grad(obj, theta),
  he = function(theta) numDeriv::hessian(obj, theta)
)

@@

ground_truth <- cbind(grid, pdf = exp(obj(grid)))

@@

pca_grid_1 <- pca_rescale(m = mu, C = cov, s = 1, k = 1)
pca_grid_3 <- pca_rescale(m = mu, C = cov, s = 1, k = 3)
pca_grid_5 <- pca_rescale(m = mu, C = cov, s = 1, k = 5)
pca_grid_7 <- pca_rescale(m = mu, C = cov, s = 1, k = 7)

plot +
  geom_point(
    data = data.frame(mvQuad::getNodes(pca_grid_7)) %>%
      mutate(weight = mvQuad::getWeights(pca_grid_7)),
    aes(x = X1, y = X2, size = weight),
    col = multi.utils::cbpalette()[2],
    inherit.aes = FALSE
  ) +
  scale_size_continuous(range = c(1, 2))

opt_bfgs <- aghq::optimize_theta(ff, mu, control = default_control(method = "BFGS"))

# Perform regular AGHQ with a product grid
norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, k = 1)
norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, k = 3)
norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, k = 5)
norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, k = 7)

# Zero is the correct answer here, so hopefully these are close!
norm_bfgs_1$lognormconst
norm_bfgs_3$lognormconst
norm_bfgs_5$lognormconst
norm_bfgs_7$lognormconst

# What about with the new PCA grid?
norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_1, adapt = FALSE)
norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_3, adapt = FALSE)
norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_5, adapt = FALSE)
norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_7, adapt = FALSE)

# These don't look near zero! Something is going wrong
norm_pca_bfgs_1$lognormconst
norm_pca_bfgs_3$lognormconst
norm_pca_bfgs_5$lognormconst
norm_pca_bfgs_7$lognormconst
```

### Other 2D function

More complicated two dimensional function from Alex's [unit tests](https://github.com/awstringer1/aghq/blob/master/tests/testthat/setup-01-optimization.R) from the `aghq` package.

@@

truemode2d <- c(log((sum(y1) + 1) / (length(y1) + 1)), log((sum(y2) + 1) / (length(y2) + 1)))
truelogint <- function(y) lgamma(1 + sum(y)) - (1 + sum(y)) * log(length(y) + 1) - sum(lgamma(y + 1))
truelognormconst2d <- truelogint(y1) + truelogint(y2)

obj <- function(x) logfteta2d(x, c(y1, y2))

ff <- list(
  fn = objfunc2d,
  gr = function(x) numDeriv::grad(obj, x),
  he = function(x) numDeriv::hessian(obj, x)
)

grid <- expand.grid(
  theta1 = seq(0, 4, length.out = 400),
  theta2 = seq(0, 4, length.out = 400)
)

ground_truth <- data.frame(grid) %>%
  rowwise() %>%
  mutate(pdf = exp(objfunc2d(x = c(theta1, theta2))))

plot <- ggplot(ground_truth, aes(x = theta1, y = theta2, z = pdf)) +
  geom_contour(col = multi.utils::cbpalette()[1]) +
  coord_fixed(xlim = c(0.8, 2), ylim = c(0.8, 2), ratio = 1) +
  labs(x = "", y = "") +
  theme_minimal()

opt_bfgs <- aghq::optimize_theta(funlist2d, c(1.5, 1.5), control = default_control(method = "BFGS"))

grid_bfgs_1 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 1)
# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
# (in rescale) no appropriate covariance matrix C
# mvQuad::rescale(grid_bfgs_1, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))

grid_bfgs_3 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 3)
# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
# (in rescale) no appropriate covariance matrix C
# mvQuad::rescale(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))

grid_bfgs_5 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 5)
# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
# (in rescale) no appropriate covariance matrix C
# mvQuad::rescale(grid_bfgs_5, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))

grid_bfgs_7 <- mvQuad::createNIGrid(dim = 2, type = "GHe", level = 7)
# Error in rescale.NIGrid(grid_bfgs_3, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian)) :
# (in rescale) no appropriate covariance matrix C
# mvQuad::rescale(grid_bfgs_7, m = opt_bfgs$mode, C = solve(opt_bfgs$hessian))

norm_bfgs_1 <- aghq::normalize_logpost(opt_bfgs, 1, 1)
norm_bfgs_3 <- aghq::normalize_logpost(opt_bfgs, 3, 1)
norm_bfgs_5 <- aghq::normalize_logpost(opt_bfgs, 5, 1)
norm_bfgs_7 <- aghq::normalize_logpost(opt_bfgs, 7, 1)

pca_grid_bfgs_1 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 1)
pca_grid_bfgs_3 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 3)
pca_grid_bfgs_5 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 5)
pca_grid_bfgs_7 <- pca_rescale(m = opt_bfgs$mode, C = solve(opt_bfgs$hessian), s = 1, k = 7)

plot_points <- function(gg) {
  plot +
    geom_point(
      data = mvQuad::getNodes(gg) %>%
        as.data.frame() %>%
        mutate(weights = mvQuad::getWeights(gg)),
      aes(x = V1, y = V2, size = weights),
      alpha = 0.8,
      col = multi.utils::cbpalette()[2],
      inherit.aes = FALSE
    ) +
    scale_size_continuous(range = c(1, 2)) +
    labs(x = "", y = "", size = "Weight") +
    theme_minimal()
}

#' AGHQ
{plot_points(grid_bfgs_1) + plot_points(grid_bfgs_3)} / {plot_points(grid_bfgs_5) + plot_points(grid_bfgs_7)}

#' PCA-AGHQ
{plot_points(pca_grid_bfgs_1) + plot_points(pca_grid_bfgs_3)} / {plot_points(pca_grid_bfgs_5) + plot_points(pca_grid_bfgs_7)}

norm_pca_bfgs_1 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_1, adapt = FALSE)
norm_pca_bfgs_3 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_3, adapt = FALSE)
norm_pca_bfgs_5 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_5, adapt = FALSE)
norm_pca_bfgs_7 <- local_normalize_logpost(opt_bfgs, basegrid = pca_grid_bfgs_7, adapt = FALSE)

results <- data.frame(
  "method" = c(
    "Truth",
    paste0("aghq, k = ", c(1, 3, 5, 7)),
    paste0("aghq-pca, k = ", c(1, 3, 5, 7))
  ),
  "lognormconst" = c(
    truelognormconst2d,
    norm_bfgs_1$lognormconst, norm_bfgs_3$lognormconst, norm_bfgs_5$lognormconst, norm_bfgs_7$lognormconst,
    norm_pca_bfgs_1$lognormconst, norm_pca_bfgs_3$lognormconst, norm_pca_bfgs_5$lognormconst, norm_pca_bfgs_7$lognormconst
  )
)

ggplot(results, aes(x = method, y = lognormconst)) +
  geom_point(size = 1) +
  labs(x = "Method", y = "Log normalising constant") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 70, vjust = 0.5, hjust = 0.5))
```

## Application to Naomi
